---
title: "Global Plastic Waste"
author: "Alex Doyle"
date: "2/15/2021"
output:
  html_document: default
  word_document: default
  pdf_document: default
---




#### Table of Contnents

1.0 Introduction

1.1 Graph 1.1 - The base R plot

1.2 Graph 1.2 - The GGplot

1.3 Graph 1.3 - The Interactive plot

1.4 Graph 1.4 - The Interactive map 

2.0 Bibliography 







#### Introduction 

Plastic waste is an ever-increasing problem on a global scale. Its effects can be seen on the world's oceans with the unfortunate existence of the Great Pacific Garbage Patch a prime example. It consists of a region of the Pacific where plastic waste is gathered by ocean currents and is estimated to cover 1.6 million km^2 (3 times the size of France)(Lebreton,  L. *et al*, 2018). The data outlined below therefore focuses heavily on the plastic waste production as well as mismanaged waste. Mismanaged waste is described as material that is littered or inadequately disposed of. This inadequately disposed waste that could eventually enter oceans via waterways, water outflows and transport by wind (Jambeck, J.R., *et al*, 2015). With the data for this analysis having been gathered for the paper "Plastic waste inputs from land into the ocean" Jambeck et a.l (2015) and made available from the [**Our World in Data**](https://ourworldindata.org/plastic-pollution) website. The paper can be found [**here**](https://science.sciencemag.org/content/347/6223/768).  

Also included required for graph 3 is the Annual Production Based emission of carbon dioxide (CO ~2~), measured in tonnes per year. It been has also been compiled by [**Out World in Data**](https://ourworldindata.org/plastic-pollution) from a number of [**sources**](https://ourworldindata.org/grapher/annual-co-emissions-by-region) and finally included was the continents data set from [**Data Hub**](https://datahub.io/) and can be found [**here**](https://datahub.io/JohnSnowLabs/country-and-continent-codes-list) and was created by [**John Snow Labs**](https://www.johnsnowlabs.com/). 





##### Data dictionary 

The data for this script consists of 3 unique files each containing over 20,000 entries. The files are separated according to 2 varibales and is outline below.



* `coast-population-vs-mismanaged-plastic.csv`


  |variable                         | class    | description                                |
  |:--------------------------------|:---------|:-------------------------------------------|
  | Entity                          | Character| Country Name                               |
  | Code                            | Character| 3 letter country code                      |
  | Year                            | Integer  | Year                                       |
  | Mismanaged plastic waste(tonned)| double   | Tonnes of mismanaged plastic waste         |
  | Coastal population              | Double   | Number of individuals living on/near coast |
  | Total population                | double   | Total population according to Gapminder    |
  
  
  

* `waste_vs_gdp`


  |variable                         | class    | description                                |
  |:--------------------------------|:---------|:-------------------------------------------|
  | Entity                          | Character| Country Name                               |
  | Code                            | Character| 3 letter country code                      |
  | Year                            | Integer  | Year                                       |
  | Per capita plastic waste(kg per person per day)| double   | Amount of plastic waste per capita in kg/day|
  | GDP per capita             | Double   | GDP per capita constant 2011 international $, rate |
  | Total population                | double   | Total population according to Gapminder    |
  
  
  

* `mismanaged_vs_gdp`


  |variable                         | class    | description                                |
  |:--------------------------------|:---------|:-------------------------------------------|
  | Entity                          | Character| Country Name                               |
  | Code                            | Character| 3 letter country code                      |
  | Year                            | Integer  | Year                                       |
  | Per capita mismanaged plastic waste(kg per day)| double   | Amount of mismanaged plastic waste per capita in kg/day|
  | GDP per capita             | Double   | GDP per capita constant 2011 international $, rate |
  | Total population                | double   | Total population according to Gapminder    | 
  
```{r libraries, include = FALSE}
library(janitor)
library(tidyverse)
library(ggplot2)
library(extrafont)
library(maps)
require(viridis)
library(plotly)
library(gapminder)
library(cowplot)
library(rgdal)
library(leaflet)
library(sp)
library(htmltools)
library(rnaturalearth)
```

```{r data, include = FALSE}
coast_vs_waste <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/coastal-population-vs-mismanaged-plastic.csv")

mismanaged_vs_gdp <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/per-capita-mismanaged-plastic-waste-vs-gdp-per-capita.csv")

waste_vs_gdp <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/per-capita-plastic-waste-vs-gdp-per-capita.csv")

continents <- readr::read_csv('https://datahub.io/JohnSnowLabs/country-and-continent-codes-list/r/0.csv') # Includes 3 letter country code and continent information for graph 1 
co2 <- readr::read_csv("https://raw.githubusercontent.com/alexdoyle115/Data_visualizations_C7083/main/Global_Plastic_Waste/co2_emissions_tonnes_per_person.csv")

```


#### Dataset sumamry {.tabset}


##### 1. Coast vs Waste
```{r summary 1, echo = FALSE}
head(coast_vs_waste)
```


##### 2. Mismanaged vs gdp
```{r summary 2, echo = FALSE}
head(mismanaged_vs_gdp)
```


##### 3. Waste vs gdp
```{r summary 3, echo = FALSE}
head(waste_vs_gdp)
```



#### Tidying Data


```{r tidy data, echo=FALSE}
# Filtering out the the data for 2010 as it is the only year with information
# Coast_vs waste data 
clean_coast <- coast_vs_waste %>%
  clean_names() %>%
  filter(year == 2010)

# mismanaged plastic waste vs gdp
clean_mismanaged <- mismanaged_vs_gdp %>%
  clean_names() %>%
  filter(year == 2010)

# plastic waste vs gdp
clean_waste <- waste_vs_gdp %>%
  clean_names() %>%
  filter(year == 2010)

er <- round(((186/20093)*100), 2)  
```



```{r removing columns, echo=FALSE}
#  dropping the duplicate rows before combining a the data sets
clean_mismanaged <- select(clean_mismanaged, -c(3, 5, 6))
clean_waste <- select(clean_waste, -c(3, 6))

clean_data <- clean_coast %>%
  inner_join(clean_mismanaged, by = c("entity", "code")) %>%
  inner_join(clean_waste, by = c("entity", "code"))

```

 
```{r renaming columns, echo=FALSE}
# renaming the columns into something much more manageablee
clean_data <- clean_data %>%
  rename(country = entity, 
         msm_plastic_per_capita = per_capita_mismanaged_plastic_waste_kilograms_per_person_per_day, 
         plastic_per_capita = per_capita_plastic_waste_kilograms_per_person_per_day, 
         msm_plastic_waste_t = mismanaged_plastic_waste_tonnes,
         gdp_per_capita =  gdp_per_capita_ppp_constant_2011_international_constant_2011_international
         ) 

```


Looking at the output the table shows that the number of rows that have data for certain variables is very low. In the `coast_vs_waste` data set had only 186 rows that were not empty working out at `r er`% of the total number of rows. On closer inspection is shows that the only year that any of the countries had data for was 2010. Cleaning up the datasets and combining them into one table with 9 rows and 235 observations. The aim of the first plot was to explore the differences in mismanaged plastic vs the more general plastic waste. The number of countries to display was far too high but by adding a dataset that contained the country codes and the continent it was possible to group the countries into the continent and look at the trends. 


#### 1.1 Graph 1.1 - The base R plot


The aim for Graph 1.1 was to compare the average amount of waste plastic and mismanaged plastic a person produces a day. While production did vary within the continents there were far too many countries to display efficiently so the countries were grouped by continents and the average was calculated. 


```{r Adding Continents, include = FALSE}

continents <- select(continents, -c(2, 3, 6)) # filtering the important columns
continents <- continents %>% # rename a columns to match the data that it will joing
  rename(code=Three_Letter_Country_Code, 
         continent = Continent_Name)

clean_data$code <- as.character(clean_data$code) # converting to character
clean_data <- clean_data %>%
  inner_join(continents, by = c("code")) # joining the data sets together

data <- clean_data %>% 
  drop_na() %>%
  filter(country != "Trinidad and Tobago") # unfortunatly Tinidad and Tobago was a massive outlier and had to removed

head(data)
```


```{r continent table, include=FALSE}
# Creating a new dataset with the 6 continents and the average plastic waste. 
tbl <- data %>% # one data set for the plastic waste
  group_by(continent) %>% 
  summarise(avg = mean(plastic_per_capita)) %>%
  mutate(
    avg = round(avg, 3), # rounding for improved readability
    continent = as.factor(continent)
  )



tbl1 <- data %>% #another one for the mismanaged plastic waste
  group_by(continent) %>%
  summarise(msmavg = mean(msm_plastic_per_capita)) %>%
  mutate(
    msmavg = round(msmavg, 3), # rounding for improved readability 
    continent = as.factor(continent)
  )


tbl <- tbl %>%
  inner_join(tbl1, by = c("continent")) # combining 2 into one for easier manouvering
tbl$continent <- as.character(tbl$continent)

tbl <- arrange(tbl, desc(avg))
```

```{r plot 1.1, echo = FALSE}

#windowsFonts("Rockwell Nova Light"  = windowsFont("Rockwell Nova Light" )) # creating the font
bg_col <- c("#E8FFFF")

palette <- c("#c88b21", "#3dac14", "#7e7fde", "#5b37ec", "#4910c9", "#AA4499") # creating a colour palette
par(mfrow = c(1, 2)) # display plots side by side
par(las=2, bg = bg_col) 



#Plot 2 
par(las=2) 
par(mar=c(4,3.5,4,2.5))

barplot(tbl$avg, 
        horiz = T,
        
        cex.names = 0.85, 
        border = "#666666",
        col = palette, 
        xlim = c(.4, 0))
text(0.4, 7.2, "Total Plastic Waste", pos = 4  )
text(0.4, 6.7, "(kg/person/day)", pos = 4  )
text(0.4, 5, "North Americans", cex = 0.8, pos = 4  )
text(0.4, 4.6, "seem to be the", cex = 0.8, pos = 4  )
text(0.4, 4.2, "runaway leader of", cex = 0.8, pos = 4  )
text(0.4, 3.8, "plastic production", cex = 0.8, pos = 4  )
# Plot 1     
barplot(tbl$msmavg,  
        horiz = T,
        xlab = "Alex Doyle | Global Plastic Waste | Graph 1", 
        cex.names = 0.8, 
        names.arg = c("North America", "South America",  "Europe      ","Asia         ", "Oceania     ", "Africa        " ),col = palette,
        xlim = c(0, .4),
        border = "#666666")

text(0.14, 7.0, "Mismanaged plastic ", pos = 4  )
text(0.14, 6.5, "Waste", pos = 4  )
text(0.14, 6.1, "(kg/person/day)", pos = 4  )
text(0.135, 2.6, "Mismanaged waste is a", pos = 4, cex = 0.8  )
text(0.135, 2.2, "material which is at high", pos = 4, cex = 0.8  )
text(0.135, 1.8, "of risk entering the", pos = 4, cex = 0.8  )
text(0.135, 1.4, "ocean", pos = 4, cex = 0.8  )
title("Daily Plastic Waste Production",outer = T, line = -2  )




```


#### 1.2 - The GGplot\
From plot one we learned that the continents that produce the most plastic waste per person per day tend to dispose of that plastic more carefully and on first appearances is seems that these higher producing countries are richer and more developed in comparison to the high mismanaged waste producers. The aim of the ggplot was therefore to look at the effects of income on the types of plastic waste produced. To make it more readable the axis are on the log scale which is important to remember. Again the plot is divided by the types of plastic and coloured according to the continents to try and decipher trends. Looking at the graph it does indicate as a countries GDP increases it seems to increase plastic waste production compared to the mismanaged waste which had a negative relationship with GDP.     

```{r, include = FALSE}

lm1 <- glm(plastic_per_capita ~ gdp_per_capita, family = binomial, data = data)
lm2 <- glm(msm_plastic_per_capita ~ gdp_per_capita, family = binomial, data = data)

summary(lm1)
summary(lm2)
```

```{r plot 1.2, echo = FALSE, message = FALSE}
# Create 2 plots side by side with ggplot

scaleFUNx <- function(x) sprintf("%.0f", x) # rounding the population GDP to 0 decimal places
scaleFUNy <- function(x) sprintf("%.3f", x) # rounding the daily plastic production to 2 decimal places


A <- ggplot(data, 
            aes(x = gdp_per_capita, y = plastic_per_capita)) +
  geom_point( # adding the points
    aes(colour = continent)) + # colour by the continents
  stat_smooth(method = "lm", formula = y ~ poly(x, 2), size = 1, se = T, colour = "black") + # adding regression lines, it is a polynomial to get better estimate on the data
  #regression line
  scale_color_manual(values = c("#c88b21", "#3dac14", "#7e7fde", "#5b37ec", "#4910c9", "#AA4499")) +
  scale_x_continuous(trans = 'log', labels=scaleFUNx) + # the rounded values to x axis
  scale_y_continuous(trans = 'log', labels=scaleFUNy) + # the rounded values to y axis
  labs(x = "Per Capita GDP (log scale)",  # labels 
       y = "Daily Waste Production (kg/per person/day) \n (log scale)",
       title = "Trends in Plastic Waste Production", 
       subtitle = "The types of plastic waste can be split a number of ways, \n one of the most important subsets of plastic waste is \n  mismanaged plastic at risk of entering the worlds oceans ") +
  theme_classic() +
  
  theme(legend.position = "none", # removing the legend from this plot
        plot.title = element_text(size  = 15, 
                                  hjust = 1), 
        plot.subtitle = element_text(size = 9, 
                                     hjust = 1), 
        plot.background = element_rect(fill = bg_col), 
        panel.background = element_rect(fill = bg_col,
                                colour = bg_col,
                                size = 0.5, linetype = "solid"))

# plot is very similar to the first except it the using the mismanaged data 
B <- ggplot(data, 
            aes(x = gdp_per_capita, y = msm_plastic_per_capita)) +
  geom_point(
    aes(colour = continent)) +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2), size = 1, se = T, colour = "black") +
  scale_color_manual(values = c("#c88b21", "#3dac14", "#7e7fde", "#5b37ec", "#4910c9", "#AA4499")) +
  scale_x_continuous(trans = 'log', labels=scaleFUNx) +
  scale_y_continuous(trans = 'log', labels=scaleFUNy) +
  labs(x = "Per Capita GDP (log scale)",
       y = "Mismanaged Plastic Waste \n (kg/person/day)",
       caption ="Alex Doyle | Global Plastic Waste | Graph 2") +
  theme_classic() +
  theme(legend.position = "bottom", # this legend will be suffice for both 
        legend.justification = c(0,1),
        legend.margin = margin(t = -15, r = 0, b = 0, l = 0, unit = "pt"),
        legend.box = "horizontal", 
        legend.title = element_blank(),
        plot.background = element_rect(fill = bg_col), 
        panel.background = element_rect(fill = bg_col,
                                colour = bg_col,
                                size = 0.5, linetype = "solid"), 
        legend.background = element_rect(fill= bg_col,
                                  size=0.5, linetype="solid", 
                                  colour = bg_col)) 

plot_grid(A, B) # plotting both plots
```



#### 1.3 Graph 1.3 - The Interactive Plot

```{r adding CO2, include = FALSE}

# new data set including the CO2 information 
co2 <- co2 %>%
  select("country" | "2010") %>% # filtering out all information relating to 2010
  rename(co2_per_capita = "2010") %>% # renaming it for the graph
  inner_join(data, by = "country") %>% # joining it to the old data set
  mutate(plastic_t = plastic_per_capita * 1000/ 365) # convert from kg/day to tonnes/year



```


```{r plot interactivity, include = FALSE}

# Create the text that will appear over the points on the interactive plot
co2 <- co2 %>%
  mutate(mytext=paste(
     country,"\n",
    "Total Population: ", total_population_gapminder, "\n", 
    "Per Capita Inforamtion:", "\n",
    "Plastic Waste (kg): ", plastic_per_capita, "\n", 
    "Mismanaged Plastic (kg): ", msm_plastic_per_capita, "\n", 
    "GDP (US$): ", round(gdp_per_capita, 1), sep=""))
```




```{r, include=FALSE}
lm <- lm(plastic_t ~ co2_per_capita, data = co2)
 summary(lm)
```



#### 1.3 - The Interactive Plot\
For the interactive plot, the aim was to add to the scatter plot in the previous as it suited the type of data that was being displayed. Adding the CO~2~ data too added another layer as it was a different way of looking at it in comparison to gdp. The interactivity also allows for zooming in on the concentrated points in the corner around the origin without needing to use the log scale. 


```{r plot 1.3, echo = FALSE, warning = FALSE, error = TRUE}
# the interactive plot 
# creating the base plot that will be made interactive
p <- co2 %>%
  ggplot() +
  geom_point(
    aes(co2_per_capita, plastic_t, size = total_population_gapminder, fill = continent, text = mytext),
    colour="black") + 
  scale_fill_manual(values = c("#c88b21", "#3dac14", "#7e7fde", "#5b37ec", "#4910c9", "#AA4499")) +
  geom_smooth( aes(co2_per_capita, plastic_t), method = "lm", se = F, colour = "black", lwd = 0.3) + 
  scale_x_continuous() +
  scale_y_continuous() +
  labs(y = "Tonnes of Plastic Produced Annually", 
       x = "Tonnes of Carbon Dioxide Produced Annually", 
       title = "Worldwide Pollution Rates") + 
  annotate(geom = "text", x = 30, y = 1.2, label = "Regression line shows \n countries polluting more or less \n than predicted in a category", size = 2.5) +
  theme_bw() +
  theme(legend.title = element_text(color = bg_col),
        legend.background = element_rect(fill= bg_col,
                                  size=0.5, linetype="solid", 
                                  colour = bg_col), 
        plot.background = element_rect(fill = bg_col), 
        panel.background = element_rect(fill = bg_col,
                                colour = bg_col,
                                size = 0.5, linetype = "solid"))
               

# convert to interactive ggplot using the ggplotly function               
ggplotly(p, tooltip = "text") %>%
  layout(margin = list(b=65,t=50), annotations = 
 list(x = 1.3, y = -0.15, text = "Alex Doyle | Global Plastice Waste | Graph 3", 
      showarrow = F, xref='paper', yref='paper', 
      xanchor='right', yanchor='auto', xshift=0, yshift=0,
      font=list(size=9)))

```




#### 1.4 - The Interactive Map\
An interactive map was a key data visualization for this assignment and this data was perfect. Colouring the map by the mismanaged plastic waste gives a good indicator of the countries that produce the most of waste. This really highlights the contribution China makes to global ocean pollution. 



```{r}
world <- (ne_countries())
shp_data <- world@data

shp_color <- data %>% 
  distinct(code, .keep_all = TRUE) %>% # removing anycountries that appear twice
  rename(adm0_a3 = code) %>% # change the 3 letter code to match the shapefile
  mutate(rank = rank(-msm_plastic_per_capita)) %>%
  select("adm0_a3", "msm_plastic_waste_t", "coastal_population", "rank")
  # the variables we want to display

shp_data <- shp_data %>%
  full_join(shp_color, by = "adm0_a3") # joining them together to before adding to the shapefile
# I found that the data reordered when I joined it directly to the shapefile therefore this keeps it in order

world <- merge(world, shp_data, by = "adm0_a3", duplicateGeoms = TRUE) # now joining to the shapefile
```

```{r}
bins <- c(0, 1000, 10000, 100000, 1000000, 5000000, 10000000) # setting the bins we want to group them into 
pal <- colorBin("YlOrRd", domain = world$msm_plastic_waste_t, bins = bins, na.color="transparent") # setting the colours
```

```{r}
tag.map.title <- tags$style(HTML(" 
  .leaflet-control.map-title { 
    transform: translate(-50%,20%);
    position: fixed !important;
    left: 50%;
    text-align: center;
    padding-left: 10px; 
    padding-right: 10px; 
    background: rgba(255,255,255,0.5);
    font-weight: bold;
    font-size: 14px;
    color: black;
  }
"))  # adding the title 

title <- tags$div(
  tag.map.title, HTML("Mismanaged Waste Producers")
)  
```



```{r}
mylabels <- paste(
  "Country: ", world@data$subunit.x, "<br/>", 
  "Mismanaged Plastic: ", world@data$msm_plastic_waste_t, " tonnes", "<br/>", 
  "Per Capita Mismanaged Waste Rank: ", world@data$rank, "<br/>", 
  world@data$economy.x, "( Scale: 1-7)", "<br/>", 
  world@data$income_grp.y, "( Scale: 1-5",  
  sep = ""
) %>%
  lapply(htmltools::HTML) # adding the text for the interactive plots 
```

```{r}

leaflet(world) %>% # adding the shapefile
  addTiles()  %>% # background map
  setView( lat=10, lng=0 , zoom=1) %>% # setting the zoom the plot will start from
  addControl(title, position = "topleft", className="map-title") %>% # adding the title to the plot
  addPolygons(
    fillColor = ~pal(msm_plastic_waste_t), # the predetermined colour palette
    stroke = TRUE, 
    fillOpacity = 0.9,
    color = 'White', 
    weight = 0.5, 
    label = mylabels,
    labelOptions = labelOptions( 
      style = list("font-weight" = "normal", padding = "3px 8px"), # label settings
      textsize = "13px", 
      direction = "auto"
    )
  ) %>%
  addLegend( pal=pal, values=~msm_plastic_waste_t, opacity=0.9, title = "Mismanaged Waste Production (t)", position = "bottomleft" ) # the legendis positioned in the bottom left

```

#### Bibliography 

| Lebreton, L., Slat, B., Ferrari, F., Sainte-Rose, B., Aitken, J., Marthouse, R.,      |   Hajbane, S., Cunsolo, S., Schwarz, A., Levivier, A. and Noble, K., 2018. Evidence   |   that the Great Pacific Garbage Patch is rapidly accumulating plastic. Scientific    |   reports, 8(1), pp.1-15.

| Jambeck, J.R., Geyer, R., Wilcox, C., Siegler, T.R., Perryman, M., Andrady, A.,       |   Narayan, R. and Law, K.L., 2015. Plastic waste inputs from land into the ocean.     |   Science, 347(6223), pp.768-771.

