---
title: "Plants in Danger"
author: "Alex Doyle"
date: "2/15/2021"
output: html_document
---
### C7083 - Data Visualization Assignment


#### Table of Contnents

1.0. Introduction

1.1. Graph 3.1 - The base R plot

1.2. Graph 3.2 - The GGplot

1.3. Graph 3.3 - The Sankey Diagram

1.4. Graph 3.4 - The Waffle Plot

2.0. Bibliography 



#### 1.0 Introduction 

The [International Union for Conservation of Natures's Red List of Threatened Species](https://www.iucnredlist.org/) is data source regarding to range, population size, conservation status and much more of all types of animal, fungi and plant species. The main aim of such a list is to provide information to government agencies, wildlife departments, NGOs and the general public about the decline in biodiversity and number of species facing extinction. The dataset for these graphics was scraped by Florent Lavergne for his [unique infographic](https://www.behance.net/gallery/98304453/Infographic-Plants-in-Danger). Florent created this project to outlines the number of plants at high risk of extinction as he felt that plants are often unerrepresented in comparission to animals at risk of extinction.  
 
```{r libraries, include = FALSE}
library(networkD3)
library(htmlwidgets)
library(htmltools)
library(tidyverse)
library(janitor)
library(GGally)
library(ggplot2)
library(ggwaffle)
library(waffle)
```
##### Data Dictionary 

* `threats.csv`

| Variable          | Class     | Description                              |
|-------------------|-----------|------------------------------------------|
| binomial_name     | character | Species name (Genus + species)           |
| country           | character | Country of origin                        |
| continent         | character | Continent of origin                      |
| group             | character | Taxonomic group                          |
| year_last_seen    | character | Period species was last seen             |
| red_list_category | character | IUCN Red List category                   |
| threat_type       | character | Type of threat                           |
| threatened        | double    | Binary 0 (not threatened) 1 (threatened) |

* `action.csv`

| Variable          | Class     | Description                              |
|-------------------|-----------|------------------------------------------|
| binomial_name     | character | Species name (Genus + species)           |
| country           | character | Country of origin                        |
| continent         | character | Continent of origin                      |
| group             | character | Taxonomic group                          |
| year_last_seen    | character | Period species was last seen             |
| red_list_category | character | IUCN Red List category                   |
| action_type       | character | Type of action                           |
| action_taken      | double    | Binary 0 (not threatened) 1 (threatened) |


```{r read in plants, include  = FALSE}
# Loading in the data 
plants <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-18/plants.csv')
actions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-18/actions.csv')
threats <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-18/threats.csv')
```
#### Tidying the Data 

The data is downloaded in 3 different sets, with the `plant.csv` being the original and the `threat.csv`and `actions.csv` are filtered down versions of it. For that reason all of the graphs are based off of these 2 smaller data sets as they were easier to explore. 

```{r, include = FALSE }
# filtering and re arranging the first plot
update <- threats %>%
  filter(threatened == 1) %>% # binary varibale to filter out the action threats 
  group_by(threat_type, year_last_seen) %>%
  summarise(
    value = sum(threatened)) %>% # count each threat for each break 
  replace_na(list(value = 0)) %>%
  spread(threat_type, value) %>%
  clean_names() %>%
  replace_na(list("geological_events" = 0, "pollution" = 0)) %>% # Replace NA's with 0s 
  ungroup() %>%
  drop_na(year_last_seen) %>% 
  mutate("year" = c("1910", "1930", "1950", "1970", "1990", "2010", "1800")) %>% # useful for reordering also tidier
  arrange(year) %>% #  reordering the rows
  rename(other = transportation_corridor) %>% # can only comfortably display 9 plots so the 4 smallest are grouped
  mutate(other = other + pollution + geological_events + human_intrusions) # creating "other

```



#### 1.1 Graph 3.1 - The base R plot

Starting off the plots it is looking at the threats that have faced plants over the decades and how they have changed. Plotting the 7 main threats, threats not recorded as well as a combination of the 3 smallest threats gives a good understanding how things have changed. Energy and Mining for example show a temporary spike that soon returns to original levels sparking interest in what caused this. 

```{r graph 3.1, echo = FALSE}
# the base R equvialnt of a facetted plot 
par(mfrow = c(3, 3), bg = "#ccffcc", mar = c(2.5, 2.5, 2.5, 2.5), oma = c(3, 0, 5, 0))

# 1. Agriculture and Aquaculture 
plot(update$year, update$agriculture_aquaculture, 
     main = "Agriculture & Aquaculture", col.main = "#003300", type = "l", ylim=range(0, 56), 
     xlab = "No", col = c("#660000"), lwd=2, 
     frame.plot = FALSE, 
     xaxt = "n")
axis(2, col="#003300", col.ticks="#003300", col.axis="#003300", cex.axis=1)
axis(1, col ="#003300", col.ticks = NA,col.axis=NA)

# Adding th text 
mtext("Threats to plants", line=1, side=3, outer=TRUE, cex=2)
mtext("Alex Doyle | Plants in Danger | Graph 1", line = 0, side = 1, outer = TRUE, cex = 0.8)

# 2. Biological Resources
plot(update$year, update$biological_resource_use, main = "Biological Resourse Use", col.main = "#003300", type = "l", ylim=range(0, 56), 
     col = c("#660000"), lwd=2, 
     frame.plot = FALSE, 
     xaxt = "n", 
     yaxt = "n")
axis(2, col ="#003300", col.ticks = NA,col.axis=NA)
axis(1, col ="#003300", col.ticks = NA,col.axis=NA)

# 3. Climate Change 
plot(update$year, update$climate_change, main = "Climate Change", col.main = "#003300", type = "l", ylim=range(0, 56), 
     col = c("#660000"), lwd=2, 
     frame.plot = FALSE, 
     xaxt = "n", 
     yaxt = "n")
axis(1, col ="#003300", col.ticks = NA,col.axis=NA)
axis(2, col ="#003300", col.ticks = NA,col.axis=NA)


# 4. Commercial development 
plot(update$year, update$commercial_development, main = "Commercial Development", col.main = "#003300", type = "l", ylim=range(0, 56), 
     col = c("#660000"), lwd=2, 
     frame.plot = FALSE, 
     xaxt = "n")
axis(2, col="#003300", col.ticks="#003300", col.axis="#003300", cex.axis=1)
axis(1, col ="#003300", col.ticks = NA,col.axis=NA)

# 5. Energy Production and Mining 
plot(update$year, update$energy_production_mining, main = "Energy and Mining", col.main = "#003300", type = "l", ylim=range(0, 56), 
     col = c("#660000"), lwd=2, 
     frame.plot = FALSE, 
     xaxt = "n", 
     yaxt = "n")
axis(1, col ="#003300", col.ticks = NA,col.axis=NA)
axis(2, col ="#003300", col.ticks = NA,col.axis=NA)

# 6. Invasive Species
plot(update$year, update$invasive_species, main = "Invasive Species", col.main = "#003300", type = "l", ylim=range(0, 56), 
     col = c("#660000"), lwd=2, 
     frame.plot = FALSE, 
     xaxt = "n", 
     yaxt = "n")
axis(1, col ="#003300", col.ticks = NA,col.axis=NA)
axis(2, col ="#003300", col.ticks = NA,col.axis=NA)

# 7. Natural System Modifications 
plot(update$year, update$natural_system_modifications, main = "Natural system changes", col.main = "#003300", type = "l", ylim=range(0, 56), 
     col = c("#660000"), lwd=2, 
     frame.plot = FALSE)
axis(1, col ="#003300", col.ticks = NA,col.axis=NA)
axis(2, col="#003300", col.ticks="#003300", col.axis="#003300", cex.axis=1)

# 8. Other factors
plot(update$year, update$other, main = "Other", col.main = "#003300", type = "l", ylim=range(0, 56), 
     col = c("#660000"), lwd=2, 
     frame.plot = FALSE,
     xaxt = "n", 
     yaxt = "n")
axis(1, col="#003300", col.ticks="#003300", col.axis="#003300", cex.axis=1)
axis(2, col="#003300", col.ticks=NA, col.axis=NA)

# 9. Unknown threats 
plot(update$year, update$unknown, main = "Unkown", col.main = "#003300", type = "l", ylim=range(0, 56), 
     col = c("#660000"), lwd=2, 
     frame.plot = FALSE,
     xaxt = "n", 
     yaxt = "n")
axis(1, col="#003300", col.ticks="#003300", col.axis="#003300", cex.axis=1)
axis(2, col="#003300", col.ticks= NA, col.axis=NA)

```




1.2 Graph 3.2 - The GGplot

The final plot that was created aims to display the change in the rate of extinction as well as the number of plants that are saved from complete extinction through intervention such as seed banks. It is simple but effective. 


```{r extinct data, include = FALSE}
# Comparing the number of extince plants to plants that have been saved
area <- actions %>%
  select(year_last_seen, red_list_category) %>%
  mutate(year_last_seen =as.factor(year_last_seen)) %>%
  mutate(year_last_seen = factor(year_last_seen, levels = c("Before 1900", "1900-1919", "1920-1939", "1940-1959", "1960-1979", "1980-1999", "2000-2020"))) %>%
  drop_na()

bg_col <- c("#E8FFFF") # saving the background colour
```

```{r graph 3.2, echo = FALSE}
# Extinct vs extinct in the wild
ggplot(area, aes(year_last_seen, fill = red_list_category)) +
  geom_bar(stat = "count") +
  scale_fill_manual(values = c("Extinct" = "#660000", "Extinct in the Wild" = "#006600")) +
  labs(x = "Year a plant was Last Seen", 
       y = "Count of plants", 
       title = "Rate of Extinction in the last 100 plus years", 
       subtitle = "The rate at which plants have gone extinct has not slowed over \n time as one would hope but as silver lining the number of plants \n that are saved from extinction through human intervention", 
       caption = "Alex Doyle | Plants in Danger | Graph 2") +
  theme_classic() +
  theme(legend.position = "top", 
        legend.box = "horizontal", 
        legend.title = element_blank(),
        plot.background = element_rect(fill = bg_col), 
        panel.background = element_rect(fill = bg_col,
                                colour = bg_col,
                                size = 0.5, linetype = "solid"), 
        legend.background = element_rect(fill= bg_col,
                                  size=0.5, linetype="solid", 
                                  colour = bg_col) )
 

```

1.3 Graph 3.3 - The Sankey Diagram
 
For the interactive plot the aim was to show the relationship between the threats, countries and how they have changed over time there were several plots that may have achieved this. The Sankey Diagram allowed for a clearer image of the "flow" of the different sections while the interactivity allows for focusing on individual nodes and streams.
```{r Sankey Diagram data, include = FALSE}
# Creating a data set for a Sankey Diagram 
part_a <- threats %>%  # filtering and mutating the data to create a to the correct shape 
  filter(threatened == 1) %>%
  group_by(year_last_seen, continent) %>%
  summarise(
    value = sum(threatened)) %>% 
  ungroup() %>%
  mutate(year_last_seen =as.factor(year_last_seen)) %>%
  mutate(year_last_seen = factor(year_last_seen, levels = c("Before 1900", "1900-1919", "1920-1939", "1940-1959", "1960-1979", "1980-1999", "2000-2020"))) %>%
  arrange(year_last_seen) %>%
  slice(-c(43:47)) %>%
  rename(source = year_last_seen, # target and source indicates the direction of the flow
         target = continent)
  
part_b <- threats %>% # as there wil be 3 levels in the diagram there will need to be new variable with one in common with the original 
  filter(threatened == 1) %>%
  group_by(continent, threat_type) %>%
  summarise(
    value = sum(threatened)) %>% 
  ungroup() %>%
  rename(target = threat_type, 
         source = continent)
  

links <- as.data.frame(rbind(part_a, part_b)) # joining the the 2 dataets 

nodes <- data.frame(name = c(as.character(links$source), # identifying the nodes
                             as.character(links$target)) %>%
                      unique())

links$IDsource <- match(links$source, nodes$name)-1 # removing frame heading and matching the sources and names
links$IDtarget <- match(links$target, nodes$name)-1 # removing frame heading and matching the targets and names
```



```{r Graph 3.3 Sankey Diagram, echo = FALSE}

# creating a Sankey Diagram 
p <- sankeyNetwork(Links = links, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "value", NodeID = "name", 
              sinksRight=FALSE, fontSize= 12, nodeWidth = 20,
              width= 700, height=550, iterations = 0) %>%
  htmlwidgets::prependContent(htmltools::tags$h1("Flow of Plant Threats"), 
                              htmltools::tags$h5("An plot highlights the major theats that have caused extinction acoss the worlds continents from the 1800s through to modern times")
                              ) # Adding some text


p <- htmlwidgets::appendContent(p, htmltools::tags$p("Alex Doyle | Plants in Danger | Graph 3"))

p

saveWidget(p, file=paste0( getwd(), "/sankeyEnergy.html"))
```

1.4 Graph 3.4 - The Waffle Plot

Across the previous plots the information has been focused on the various threats to the plants that may have led to the plants going extinct (in the wild). The actions data set received little attention; this was mainly due to its contents. Looking at the waffle plot below it clearly shows the number of unknown/no actions taken outweighs all the other actions. The waffle plot offers the ability to show the imbalance and highlights the lack of Education & Awareness actions being taken.  

```{r waffle data, include = FALSE}

# The waffle plot 
waffle1 <- actions %>% 
  filter(action_taken == 1) %>%
  group_by(action_type, year_last_seen) %>%
  summarise(
    value = sum(action_taken))


parts <- c(`Education & Awareness` = 1 ,`Land & Water Protection` = 65, `Law & Policy` = 29, 
           `Research & Moitoring` = 27, `Species Management` = 58, `Unknown` =379)

focus <- c(`Education & Awareness` = 1 ,`Land & Water Protection` = 65, `Law & Policy` = 29, 
           `Research & Moitoring` = 27, `Species Management` = 58)

```


```{r graph 3.4, echo = FALSE}
waffle(parts, rows=20, size=1, 
       title="Actions Taken to Prevent Extinction",
       xlab="One square = One Action Taken                    Alex Doyle | Plants in Danger | Graph 4")

```

2.0 Bibliography


